Stochastic trace estimation for parameter-dependent matrices applied to spectral density approximation

### 2025-02-25 paper/analysis.tex
address review comments, clean up equations, create n_c=2 plot

We first derive an $L^1$-error bound for the Girard-Hutchinson estimator~\refequ{equ:hutchinson-trace-estimator} that is relative to the $L^1$-norms of the Frobenius and spectral norms of $\mtx{B}(t)$, defined by
\[
 \|\mtx{B}(t)\|_{F,1} := \int_{a}^{b} \|\mtx{B}(t)\|_{F}~\mathrm{d}t, \quad  \|\mtx{B}(t)\|_{2,1} := \int_{a}^{b} \|\mtx{B}(t)\|_{2}~\mathrm{d}t.
\]

%\begin{theorem}{Girard-Hutchinson estimator for parameter-dependent matrices}{hutchinson}
%    For a symmetric matrix $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ that depends continuously on $t \in [a, b]$, consider the  Girard-Hutchinson estimator~\refequ{equ:hutchinson-trace-estimator} with an $n\times n_{\mtx{\Psi}}$ Gaussian random matrix $\mtx{\Psi}$. Then for any $p \geq 1, p \in \mathbb{R}$ and $\gamma \geq 1$ the bound 
%    \begin{equation} \label{eq:ghbound}
%        \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t \leq c \gamma \max\left\{ \sqrt{\frac{p}{n_{\mtx{\Psi}}}}, \frac{p}{n_{\mtx{\Psi}}}  \right\} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
%    \end{equation}
%    holds with probability at least $1 - \gamma^{-p}$.
%    In particular, given $\varepsilon \in (0, ce)$ and $\delta \in (0, e^{-1})$, the bound $\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) | ~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t$ holds 
%    with probability at least $1-\delta$ if $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1}))$.
%\end{theorem}

\begin{theorem}{Girard-Hutchinson estimator for parameter-dependent matrices}{hutchinson}
    For a symmetric matrix $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ that depends continuously on $t \in [a, b]$, consider the  Girard-Hutchinson estimator~\refequ{equ:hutchinson-trace-estimator} with an $n\times n_{\mtx{\Psi}}$ Gaussian random matrix $\mtx{\Psi}$. Then for any $p \geq 1$, $p \in \mathbb{R}$ and $\gamma \geq 1$, the following bound holds with probability at least $1 - \gamma^{-p}$:
    \begin{equation} \label{eq:ghbound}
        \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t \leq 4 \gamma \Big( \sqrt{\frac{p}{n_{\mtx{\Psi}}}}  \|\mtx{B}(t)\|_{F,1} + \frac{2 p}{n_{\mtx{\Psi}}} \|\mtx{B}(t)\|_{2,1} \Big).
    \end{equation}
In particular, given $\varepsilon \in (0, 1)$ and $\delta \in (0, e^{-1}]$, the bound $\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) | ~\mathrm{d}t \leq \varepsilon \|\mtx{B}(t)\|_{F,1}$ holds 
    with probability at least $1-\delta$ if $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1}))$.
\end{theorem}
\begin{proof}
    Let us first consider $t$ fixed. From the proof of \cite[Theorem 1]{cortinovis-2022-randomized-trace}, we know that $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$ is a $(2 \lVert \mtx{B}(t) \rVert _F^2 / n_{\mtx{\Psi}}, 2 \lVert \mtx{B}(t) \rVert _2 / n_{\mtx{\Psi}})$-sub-gamma random variable. Therefore, \reflem{lem:sub-gamma-moments} yields
    \begin{equation}
        \mathbb{E}^{p}[\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))] \leq 4  \sqrt{\frac{p}{n_{\mtx{\Psi}}}} \lVert \mtx{B}(t) \rVert _F + 8 \frac{p}{n_{\mtx{\Psi}}} \lVert \mtx{B}(t) \rVert _2.
        \label{equ:moment-bound-hutchinson}
    \end{equation}
%    The proof proceeds by bounding moments for fixed values of $t$ and then uses Minkowski's integral inequality. We start with a standard embedding trick (see, e.g.,~\cite[Theorem 1]{cortinovis-2022-randomized-trace}) for which we define
%    \begin{equation}
%        \widetilde{\mtx{B}}(t)
%        = \frac{1}{n_{\mtx{\Psi}}} \begin{pmatrix}
%            \mtx{B}(t) & & \\
%            & \ddots & \\
%            & & \mtx{B}(t)
%        \end{pmatrix}
%        \quad \text{and} \quad
%        \widetilde{\vct{\psi}} = \begin{pmatrix}
%            \vct{\psi}_1 \\
%            \vdots \\
%            \vct{\psi}_{n_{\mtx{\Psi}}}
%        \end{pmatrix}.
%    \end{equation}
%    Note that $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) =  \Trace(\widetilde{\mtx{B}}(t)) - \widetilde{\vct{\psi}}^{\top} \widetilde{\mtx{B}}(t) \widetilde{\vct{\psi}}$, $\lVert \widetilde{\mtx{B}}(t) \rVert _2 = \lVert \mtx{B}(t) \rVert _2 / n_{\mtx{\Psi}}$ and $\lVert \widetilde{\mtx{B}}(t) \rVert _F = \lVert \mtx{B}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$ hold. Therefore, by \reflem{lem:hutchinson-sub-exponential}, $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$ is a centered and $(2 \lVert \mtx{B}(t) \rVert _F^2 (1 - 1/\beta)^{-1} / n_{\mtx{\Psi}}, 2 \lVert \mtx{B}(t) \rVert _2 \beta / n_{\mtx{\Psi}})$-sub-exponential random variable for any $\beta > 1$. Thus, we can apply \reflem{lem:sub-exponential-moments} to bound
%    \begin{equation}
%        \mathbb{E}^{p}[\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))] \leq 2 \left(\sqrt{\frac{\pi}{1-1/\beta}} \sqrt{\frac{p}{n_{\mtx{\Psi}}}} \lVert \mtx{B}(t) \rVert _F + 4 \beta \frac{p}{n_{\mtx{\Psi}}} \lVert \mtx{B}(t) \rVert _2 \right).
%    \end{equation}
%    Bounding $\lVert \mtx{B}(t) \rVert _2 \leq \lVert \mtx{B}(t) \rVert _F$ and letting $\beta = (1 + \sqrt{1 + \pi/4}) / 2$ we get
%    \begin{equation}
%        \mathbb{E}^{p}[\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))] \leq \underbrace{4 (1 + \sqrt{1 + \pi/4})}_{=~c} \left( \sqrt{\frac{p}{n_{\mtx{\Psi}}}} + \frac{p}{n_{\mtx{\Psi}}} \right) \lVert \mtx{B}(t) \rVert _F.
%        \label{equ:moment-bound-hutchinson}
%    \end{equation}

    To address the parameter-dependent case, we first note that the continuity assumption implies that $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$ is measurable. Therefore, we can apply Minkowski's integral inequality~\cite[Theorem 202]{hardy-1952-inequalities} to conclude from the moment bound~\refequ{equ:moment-bound-hutchinson} that
    \begin{align}
        \mathbb{E}^{p}\left[ \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t  \right]
        &\leq \int_{a}^{b} \mathbb{E}^{p}\left[ \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) \right]~\mathrm{d}t \notag \\
        &\leq 4 \sqrt{\frac{p}{n_{\mtx{\Psi}}}} \lVert \mtx{B}(t) \rVert_{F,1} + 8 \frac{p}{n_{\mtx{\Psi}}} \lVert \mtx{B}(t) \rVert_{2,1}.
    \end{align}
    This implies~\refequ{eq:ghbound} by Markov's inequality.

@@

%    we note that $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) = r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}})$ and $\lVert \widetilde{\mtx{B}}(t) \rVert _F = \lVert \mtx{B}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$ hold. Thus, the claimed probabilistic bound~\refequ{eq:ghbound} follows from applying~\refequ{equ:hutchinson-trace-onequery-uniform} to $r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}})$.

@@

\begin{lemma}{Nyström approximation for parameter-dependent matrices}{nystrom}
    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be SPSD and continuous for $t \in [a, b]$. Consider the Nyström approximation $\Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ with a Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and \emph{even} $n_{\mtx{\Omega}} \geq 4$.
    Then, for any $\gamma \geq 1$, the bound
    \begin{equation}
        \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t
        \label{qu:nystrompp-theorem-bound}
    \end{equation}
        holds with probability at least $1 - \gamma^{-n_{\mtx{\Omega}} / 4}$ for 
        $c = 154$. In particular, given $\varepsilon > 0$ and $\delta \in (0, 1)$, the bound $\int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t$ holds with probability at least $1-\delta$ if $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2} + \log(\delta^{-1}))$.
\end{lemma}

@@

    \[

@@

    \]

@@

    \begin{equation} \label{equ:pinv-spectral-norm-bound}

@@

@@

@@

@@

    Inserting~\refequ{equ:nystrom-proof-tail-bound} along with \refequ{equ:nystrom-proof-frobenius-trace} in \refequ{equ:nystrom-proof-persson-bonud}, letting $c=154$, and using the triangle inequality for $\mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}[\cdot]$, we obtain
    \begin{equation} \label{eq:blubber}
        \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}} \left[\lVert \mtx{B} - \Nystr{\mtx{\Omega}}{\mtx{B}} \rVert _F \right]
        \leq \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}} \left[ \lVert \mtx{\Lambda}_2 \rVert _F \right] + \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}} \left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
        \leq \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \Trace(\mtx{B}).
    \end{equation}
    As in \cite{kressner-2024-randomized-lowrank}, one can show that the error $\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F$ is measurable.
    As in the proof of~\refthm{thm:hutchinson}, the claimed bound~\refequ{qu:nystrompp-theorem-bound}
    follows from~\refequ{eq:blubber} using Minkowski's integral inequality and Markov's inequality.
%     
%     
%         
%     and by the Minkowski's integral inequality \cite[Theorem 202]{hardy-1952-inequalities}, we get for $n_{\mtx{\Omega}} \geq 4$
%     \begin{align}
%         \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \right] 
%         &\leq \int_{a}^{b} \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}} \left[\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F \right] ~\mathrm{d}t \notag \\
%         %&\leq \mathbb{E}^{\sfrac{k}{2}} \left[ \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t)) + \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] && \text{(\refequ{xy})} \notag \\
%         %&\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t + \mathbb{E}^{\sfrac{k}{2}} \left[ \int_{a}^{b}  \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] \notag \\
%         %&\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t + \sqrt{k} \int_{a}^{b} \mathbb{E}^{\sfrac{k}{2}} \left[ \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2 \right]  ~\mathrm{d}t \notag \\
%         &\leq \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t.
%     \end{align}
%     From Markov's inequality follows with probability at least $1 - \gamma^{-\sfrac{n_{\mtx{\Omega}}}{4}}$
%     \begin{equation}
%         \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t,
%     \end{equation}
%     from which follows the first statement.

>>>>

We first derive an $L^1$-error bound for the Girard-Hutchinson estimator~\refequ{equ:hutchinson-trace-estimator} that is relative to the $L^1$-norms of the Frobenius and spectral norms of $\mtx{B}(t)$, defined by
\begin{equation*}
 \|\mtx{B}(t)\|_{F,1} := \int_{a}^{b} \|\mtx{B}(t)\|_{F}~\mathrm{d}t, \quad  \|\mtx{B}(t)\|_{2,1} := \int_{a}^{b} \|\mtx{B}(t)\|_{2}~\mathrm{d}t.
\end{equation*}

%\begin{theorem}{Girard-Hutchinson estimator for parameter-dependent matrices}{hutchinson}
%    For a symmetric matrix $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ that depends continuously on $t \in [a, b]$, consider the  Girard-Hutchinson estimator~\refequ{equ:hutchinson-trace-estimator} with an $n\times n_{\mtx{\Psi}}$ Gaussian random matrix $\mtx{\Psi}$. Then for any $p \geq 1, p \in \mathbb{R}$ and $\gamma \geq 1$ the bound 
%    \begin{equation} \label{equ:girard-hutchinson-bound}
%        \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t \leq c \gamma \max\left\{ \sqrt{\frac{p}{n_{\mtx{\Psi}}}}, \frac{p}{n_{\mtx{\Psi}}}  \right\} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
%    \end{equation}
%    holds with probability at least $1 - \gamma^{-p}$.
%    In particular, given $\varepsilon \in (0, ce)$ and $\delta \in (0, e^{-1})$, the bound $\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) | ~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t$ holds 
%    with probability at least $1-\delta$ if $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1}))$.
%\end{theorem}

\begin{theorem}{Girard-Hutchinson estimator for parameter-dependent matrices}{hutchinson}
For a symmetric matrix $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ that depends continuously on $t \in [a, b]$, consider the  Girard-Hutchinson estimator~\refequ{equ:hutchinson-trace-estimator} with an $n\times n_{\mtx{\Psi}}$ Gaussian random matrix $\mtx{\Psi}$. Then for any $p \geq 1$, $p \in \mathbb{R}$ and $\gamma \geq 1$, the following bound holds with probability at least $1 - \gamma^{-p}$:
\begin{equation}
    \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t \leq 4 \gamma \Big( \sqrt{\frac{p}{n_{\mtx{\Psi}}}}  \|\mtx{B}(t)\|_{F,1} + \frac{2 p}{n_{\mtx{\Psi}}} \|\mtx{B}(t)\|_{2,1} \Big).
    \label{equ:girard-hutchinson-bound}
\end{equation}
In particular, given $\varepsilon \in (0, 1)$ and $\delta \in (0, e^{-1}]$, the bound $\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) | ~\mathrm{d}t \leq \varepsilon \|\mtx{B}(t)\|_{F,1}$ holds 
    with probability at least $1-\delta$ if $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1}))$.
\end{theorem}
\begin{proof}
    Let us first consider $t$ fixed. From the proof of \cite[Theorem 1]{cortinovis-2022-randomized-trace}, we know that $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$ is a $(2 \lVert \mtx{B}(t) \rVert _F^2 / n_{\mtx{\Psi}}, 2 \lVert \mtx{B}(t) \rVert _2 / n_{\mtx{\Psi}})$-sub-gamma random variable. Therefore, \reflem{lem:sub-gamma-moments} yields
    \begin{equation}
        \mathbb{E}^{p}[\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))] \leq 4  \sqrt{\frac{p}{n_{\mtx{\Psi}}}} \lVert \mtx{B}(t) \rVert _F + 8 \frac{p}{n_{\mtx{\Psi}}} \lVert \mtx{B}(t) \rVert _2.
        \label{equ:moment-bound-hutchinson}
    \end{equation}
%    The proof proceeds by bounding moments for fixed values of $t$ and then uses Minkowski's integral inequality. We start with a standard embedding trick (see, e.g.,~\cite[Theorem 1]{cortinovis-2022-randomized-trace}) for which we define
%    \begin{equation}
%        \widetilde{\mtx{B}}(t)
%        = \frac{1}{n_{\mtx{\Psi}}} \begin{pmatrix}
%            \mtx{B}(t) & & \\
%            & \ddots & \\
%            & & \mtx{B}(t)
%        \end{pmatrix}
%        \quad \text{and} \quad
%        \widetilde{\vct{\psi}} = \begin{pmatrix}
%            \vct{\psi}_1 \\
%            \vdots \\
%            \vct{\psi}_{n_{\mtx{\Psi}}}
%        \end{pmatrix}.
%    \end{equation}
%    Note that $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) =  \Trace(\widetilde{\mtx{B}}(t)) - \widetilde{\vct{\psi}}^{\top} \widetilde{\mtx{B}}(t) \widetilde{\vct{\psi}}$, $\lVert \widetilde{\mtx{B}}(t) \rVert _2 = \lVert \mtx{B}(t) \rVert _2 / n_{\mtx{\Psi}}$ and $\lVert \widetilde{\mtx{B}}(t) \rVert _F = \lVert \mtx{B}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$ hold. Therefore, by \reflem{lem:hutchinson-sub-exponential}, $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$ is a centered and $(2 \lVert \mtx{B}(t) \rVert _F^2 (1 - 1/\beta)^{-1} / n_{\mtx{\Psi}}, 2 \lVert \mtx{B}(t) \rVert _2 \beta / n_{\mtx{\Psi}})$-sub-exponential random variable for any $\beta > 1$. Thus, we can apply \reflem{lem:sub-exponential-moments} to bound
%    \begin{equation}
%        \mathbb{E}^{p}[\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))] \leq 2 \left(\sqrt{\frac{\pi}{1-1/\beta}} \sqrt{\frac{p}{n_{\mtx{\Psi}}}} \lVert \mtx{B}(t) \rVert _F + 4 \beta \frac{p}{n_{\mtx{\Psi}}} \lVert \mtx{B}(t) \rVert _2 \right).
%    \end{equation}
%    Bounding $\lVert \mtx{B}(t) \rVert _2 \leq \lVert \mtx{B}(t) \rVert _F$ and letting $\beta = (1 + \sqrt{1 + \pi/4}) / 2$ we get
%    \begin{equation}
%        \mathbb{E}^{p}[\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))] \leq \underbrace{4 (1 + \sqrt{1 + \pi/4})}_{=~c} \left( \sqrt{\frac{p}{n_{\mtx{\Psi}}}} + \frac{p}{n_{\mtx{\Psi}}} \right) \lVert \mtx{B}(t) \rVert _F.
%        \label{equ:moment-bound-hutchinson}
%    \end{equation}

    To address the parameter-dependent case, we first note that the continuity assumption implies that $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$ is measurable. Therefore, we can apply Minkowski's integral inequality~\cite[Theorem 202]{hardy-1952-inequalities} to conclude from the moment bound~\refequ{equ:moment-bound-hutchinson} that
    \begin{align*}
        \mathbb{E}^{p}\left[ \int_{a}^{b} |\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))|~\mathrm{d}t  \right]
        &\leq \int_{a}^{b} \mathbb{E}^{p}\left[ \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) \right]~\mathrm{d}t \notag \\
        &\leq 4 \sqrt{\frac{p}{n_{\mtx{\Psi}}}} \lVert \mtx{B}(t) \rVert_{F,1} + 8 \frac{p}{n_{\mtx{\Psi}}} \lVert \mtx{B}(t) \rVert_{2,1}.
    \end{align*}
    This implies~\refequ{equ:girard-hutchinson-bound} by Markov's inequality.

@@

%    we note that $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) = r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}})$ and $\lVert \widetilde{\mtx{B}}(t) \rVert _F = \lVert \mtx{B}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$ hold. Thus, the claimed probabilistic bound~\refequ{equ:girard-hutchinson-bound} follows from applying~\refequ{equ:hutchinson-trace-onequery-uniform} to $r(\widetilde{\mtx{B}}(t), \widetilde{\vct{\psi}})$.

@@

\begin{lemma}{Nyström approximation for parameter-dependent matrices}{nystrom}
    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be SPSD and continuous for $t \in [a, b]$. Consider the Nyström approximation $\Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ with a Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and \emph{even} $n_{\mtx{\Omega}} \geq 4$.
    Then, for any $\gamma \geq 1$, the bound
    \begin{equation}
        \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t
        \label{equ:nystrompp-theorem-bound}
    \end{equation}
        holds with probability at least $1 - \gamma^{-n_{\mtx{\Omega}} / 4}$ for 
        $c = 154$. In particular, given $\varepsilon > 0$ and $\delta \in (0, 1)$, the bound $\int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t$ holds with probability at least $1-\delta$ if $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2} + \log(\delta^{-1}))$.
\end{lemma}

@@

    \begin{equation*}

@@

    \end{equation*}

@@

@@

        \label{equ:pinv-spectral-norm-bound}

@@

    \begin{equation*}

@@

    \end{equation*}

@@

    Inserting~\refequ{equ:nystrom-proof-tail-bound} along with \refequ{equ:nystrom-proof-frobenius-trace} in \refequ{equ:nystrom-proof-persson-bonud}, letting $c=154$, and using the triangle inequality for $\mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}[\cdot]$, we obtain
    \begin{equation}
        \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}} \left[\lVert \mtx{B} - \Nystr{\mtx{\Omega}}{\mtx{B}} \rVert _F \right]
        \leq \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}} \left[ \lVert \mtx{\Lambda}_2 \rVert _F \right] + \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}} \left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
        \leq \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \Trace(\mtx{B}).
        \label{equ:frobenius-moment-bound}
    \end{equation}
    As in \cite{kressner-2024-randomized-lowrank}, one can show that the error $\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F$ is measurable.
    As in the proof of~\refthm{thm:hutchinson}, the claimed bound~\refequ{equ:nystrompp-theorem-bound}
    follows from~\refequ{equ:frobenius-moment-bound} using Minkowski's integral inequality and Markov's inequality.
%     
%     
%         
%     and by the Minkowski's integral inequality \cite[Theorem 202]{hardy-1952-inequalities}, we get for $n_{\mtx{\Omega}} \geq 4$
%     \begin{align}
%         \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \right] 
%         &\leq \int_{a}^{b} \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}} \left[\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F \right] ~\mathrm{d}t \notag \\
%         %&\leq \mathbb{E}^{\sfrac{k}{2}} \left[ \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t)) + \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] && \text{(\refequ{xy})} \notag \\
%         %&\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t + \mathbb{E}^{\sfrac{k}{2}} \left[ \int_{a}^{b}  \sqrt{k} \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2~\mathrm{d}t  \right] \notag \\
%         %&\leq \frac{1}{\sqrt{k}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t + \sqrt{k} \int_{a}^{b} \mathbb{E}^{\sfrac{k}{2}} \left[ \lVert \mtx{\Lambda}_2(t)^{\sfrac{1}{2}} \mtx{\Omega}_2(t) \mtx{\Omega}_1(t)^{\dagger} \rVert _2^2 \right]  ~\mathrm{d}t \notag \\
%         &\leq \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t.
%     \end{align}
%     From Markov's inequality follows with probability at least $1 - \gamma^{-\sfrac{n_{\mtx{\Omega}}}{4}}$
%     \begin{equation}
%         \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t,
%     \end{equation}
%     from which follows the first statement.
